Laurentia's continued motion into the Neoproterozoic: new paleomagnetic pole from the Jacobsville sandstone

### 2022-03-25 Manuscript/Manuscript.tex

The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and the associated deposition of sediments in the interior of Laurentia while the continent traveled rapidly from high to low latitudes. Coeval with the cessation of rifting and decrease in the speed of plate motion is the ca. 1090-980 Ma Grenvillian Orogeny whose far-field stress is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation. A lack of paleomagnetic record from the interior of Laurentia between the end of the Midcontinent Rift and the cooling of rocks associated with the Grenvillian Orogeny resulted in a gap of record for the position and orientation of Laurentia for nearly 100 Myr. To fill this gap, we develop a new paleomagnetic pole from from red beds of the Jacobsville Formation along 5 stratigraphic sections in Northern Michigan, US. High-resolution thermal demagnetization experiments together with successful fold tests resolve primary magnetic remanence which show shallow inclinations despite being  flattened. Inclination shallowing corrected paleomagnetic pole position places Laurentia at low latitudes south of the Keweenawan Track ca. 990 Ma, confirming continued plate motion of Laurentia across the equator in the Late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the Grenville Loop poles are younger than previously thought.

>>>>

The ca. 1.1 Ga North American Midcontinent Rift resulted in protracted magmatism and the associated deposition of sediments in the interior of Laurentia while the continent traveled rapidly from high to low latitudes. Coeval with the cessation of rifting and decreasing in the speed of plate motion is the ca. 1090-980 Ma Grenvillian Orogeny whose far-field stress is interpreted to have propagated into the interior of Laurentia and inverted the Midcontinent Rift, deforming the post-rift Jacobsville Formation. A lack of paleomagnetic record from the interior of Laurentia between the end of the Midcontinent Rift and the cooling of rocks associated with the Grenvillian Orogeny resulted in poorly constrained position and orientation of Laurentia for nearly 100 Myr. To fill this gap, we develop a new paleomagnetic pole from red beds of the Jacobsville Formation along 5 stratigraphic sections in Northern Michigan, US. Our high-resolution thermal demagnetization experiments successfully resolve primary magnetic remanence that pass a fold test and a conglomerate test. Our results agree with previous chemical demagnetization data but provide additional constraints on the shallowing of the inclinations and on the age of the pole once paired with max deposition ages developed from high-precision detrital zircon dating. Inclination shallowing corrected paleomagnetic pole position places Laurentia at low latitudes south of the Keweenawan Track ca. 990 Ma, confirming continued plate motion of Laurentia across the equator in the Late Mesoproterozoic to early Neoproterozoic. The new pole from the interior of Laurentia indicates the Grenville Loop poles are younger than previously thought.

\begin{keypoints}
\item Laurentia  
\item Inclination shallowing  
\item Paleogeography  
\item Grenville  
\item Keweenawan Superchron  
\item Red Bed 
\end{keypoints}
